The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile

### 2024-03-20 article/paper_drought_Chile2023.tex
update article

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    drought \sep land cover change \sep 
    satellite
\end{keyword}
\end{frontmatter}
    \captionsetup{justification=raggedright,singlelinecheck=false}

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, breakable, sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

For monitoring drought, the World Meteorological Organization recommends
the SPI (Standardized Precipitation Index) \citep{WMO2012}. The SPI is a
multi-scalar drought index that only uses precipitation to assess short-
to long-term droughts. The primary cause of drought is precipitation
anomalies, and temperature usually makes it worse \citep{Luo2017}.
Nowadays, there is an increase in attention toward using AED separately
to monitor droughts \citep{Vicente-Serrano2020}. One reason is due to
its attribution to increasing flash droughts in water-limited regions
\citep{Noguera2022}. \citet{Vicente-Serrano2010} proposed the
Standardized Precipitation Evapotranspiration Index (SPEI), which
incorporated the temperature effect by subtracting AED from
precipitation. SPEI allows for analysis of the combined effect of
precipitation and AED. Since its formulation, it has been used worldwide
for the study and monitoring of drought
\citep{Gebrechorkos2023, Liu2021}. \citet{Hobbins2016} and
\citet{McEvoy2016} developed the Evaporative Demand Drought Index (EDDI)
to monitor droughts solely using the AED, and it has proven effective in
monitoring flash droughts \citep{Li2024, Ford2023}. For soil moisture,
several drought indices exist, such as the Soil Moisture Deficit Index
(SDMI) \citep{Narasimhan2005} and the Soil Moisture Agricultural Drought
Index (SMADI) \citep{Souza2021}. \citet{Hao2013} and
\citet{AghaKouchak2014} proposed the Standardized Soil Moisture Index
(SSI), which has a similar formulation as the SPI, SPEI, and EDDI. Thus,
there are plenty of drought indices that allow for a comprehensive
assessment of drought from short- to long-term scales and consider
variables from the earth's water balance on their own (e.g.,
precipitation, AED, soil moisture). Using this information, we can
advance our understanding of the impact of drought on ecosystems.

Chile's diverse climatic and ecosystem types
\citep{Beck2023, Luebert2022} make it an ideal natural laboratory for
studying climate and ecosystems. Additionally, the country has
experienced severe drought conditions that have had significant effects
on vegetation and water storage. North-central Chile has faced a
persistent precipitation deficit since 2010, defined as a mega drought.
\citep{Garreaud2017}, which has impacted the Chilean ecosystem. This
megadrought was defined by the Standardized Precipitation Index (SPI) of
twelve months in December having values below one standard deviation.
Some studies have addressed how this drought affects single ecosystems
in terms of forest development \citep{Miranda2020, Venegas2018}, forest
fire occurrence \citep{UrrutiaJalabert2018}, and crop productivity
\citep{Zambrano2023, Zambrano2018, Zambrano2016}. We found one study
regarding land cover and drought in Chile. The study by
\citet{Fuentes2021} evaluates water scarcity and land cover change in
Chile between 29° and 39° of south latitude. \citet{Fuentes2021} used
the SPEI of one month for evaluating drought, which led to misleading
results. For example, they did not find a temporal trend in the SPEI but
found a decreasing trend in water availability and an increase trend on
AED, which in turn should have been capable of being captured with
longer time scales of the SPEI. The term ``megadrought'' in Chile is
used to describe a prolonged water shortage that lasts for several
years, resulting in a permanent deficit that impacts the hydrological
system \citep{Boisier2018}. Therefore, it is crucial to evaluate
temporal scales that consider the cumulative impact over a period of
several years. There is limited understanding of the correlation between
drought and the environment in Chile. Therefore, it is crucial to have a
deeper understanding of how meteorological and soil moisture droughts
impact the dynamics of the environment in order to make informed
decisions on how to adapt.

@@

\hypertarget{trend-of-short--to-long-term-drought}{%
\subsection{Trend of short- to long-term
drought}\label{trend-of-short--to-long-term-drought}}

@@

We analyzed the trend of vegetation productivity over the unchanged land
cover macroclasses. To achieve this, we used the persistent mask of land
cover macroclasses. This way, we tried to reduce the noise in the
vegetation due to a change in land cover from year to year.

We examine the drought indices of water demand, water supply, and soil
moisture and their correlation with vegetation productivity. The
objective is to determine the impact of soil moisture and water demand
and supply on vegetation productivity. We want to address three main
questions: Which of the drought variables---supply, demand, or soil
moisture---most helps to explain the changes in plant productivity?
Which of the short-term or long-term time scales of the drought variable
has a bigger effect on vegetation productivity in Chile? And finally,
how strong is the relationship between the variables and the drought
index? Thus, we will be able to advance in understanding how climate is
affecting vegetation, considering the impact on the five land cover
types: forest, cropland, grassland, savanna, and shrubland.

We conducted an analysis on the linear correlation between the indices
SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36
months with zcNDVI. We used a method similar to that used by
\citet{Meroni2017} which compared the SPI with the cumulative FAPAR
(Fraction of Absorbed Photosynthetically Active Radiation). A
pixel-to-pixel linear correlation analysis was performed for each index
within the persistent mask of land cover macroclasses. To begin, the
Pearson coefficient of correlation is computed for each of the six time
scales. A time scale is identified as the one that attains the highest
correlation (p \textless{} 0.05). Subsequently, the Pearson correlation
coefficient corresponding to the time scales at which the value peaked
was extracted. As a result, for each index, we generated two raster
maps: 1) containing the raster with values of the time scales and
drought index that reached the maximum correlation, and 2) having the
magnitude of the correlation obtained by the drought index at the time
scales.

@@

For the downloading, processing, and analysis of the spatio-temporal
data, we used the open source software for statistical computing and
graphics, \texttt{R} \citep{R2023}. For downloading ERA5L, we used the
\texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing raster
data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
used \texttt{\{SPEI\}} \citep{Bergueria2023}. For mapping, we use
\{tmap\} \citep{Tennekes2018}. For data analysis, the suite
\{tidyverse\} \citep{Wickham2019} was used. For the random forest
modeling, we used the \{tidymodels\}\citep{Kuhn2020} and
\{ranger\}\citep{Wright2017} packages.

@@

\hypertarget{trend-of-short--to-long-term-drought-1}{%
\subsection{Trend of short- to long-term
drought}\label{trend-of-short--to-long-term-drought-1}}

@@

In Figure~\ref{fig-trendDIMacro}, the averaged aggregation per
macrozone, drought index, and time scale are shown. The macrozones that
have the lowest trend are ``Norte Chico'' and ``Centro,'' where the SPI,
SPEI, and SSI show that it decreases at longer time scales. Potentially
explained due to the prolonged reduction in precipitation that has
affected the hydrological system in Chile. At 36 months, it reaches
trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and
SSI. For ``Sur,'' the behavior is similar, decreasing at longer scales
and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI.
``Norte Grande'' has the highest trend at 36 months for EDDI (0.042 per
decade), and ``Centro'' has the lowest for SPI and SPEI. In ``Norte
Grande'' and ``Norte Chico,'' which are in a semi-arid climate, it is
evident that the EDDI has an effect on the difference between the SPI
and SPEI index, which is not seen in the other macrozones. Contrary to
the other macrozones, ``Austral'' showed an increase in all indices,
being the highest for EDDI at 36 months (0.025) and the lowest for SSI,
which shows only a minor increase in the trend.

\hypertarget{interaction-of-land-cover-and-drought-1}{%
\subsubsection{Interaction of land cover and
drought}\label{interaction-of-land-cover-and-drought-1}}

\hypertarget{land-cover-change-1}{%
\subsection{Land cover change}\label{land-cover-change-1}}

@@

From the trend analysis in Table \ref{tab-landcoverTrend}, we can
indicate that the ``Norte Chico'' shows an increase in barren land of
111 \(km^2 yr^{-1}\) and a reduction in the class savanna of 70
\(km^2 yr^{-1}\). In the ``Centro'' and ``Sur,'' there are changes with
an important reduction in savanna (136 to 318 \(km^2 yr^{-1}\) ), and an
increase in shrubland and grassland. Showing a change for more dense
vegetation types. The area under cultivation (croplands) appears to be
shifting from the ``Centro'' to the ``Sur.'' Also, there is a high
increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' replacing the
savanna lost (Table \ref{tab-landcoverTrend}).

@@

In Figure~\ref{fig-zcNDVI_var} it is showed the spatial map of trends in
zcNDVI (Figure~\ref{fig-zcNDVI_var}a). In ``Norte Grande,'' vegetation
productivity, as per the z-index, exhibits a yearly increase of 0.027
for grassland and 0.032 for shrubland. In the ``Norte Chico,'' savanna
has the lowest trend of -0.062, cropland -0.047, shrubland -0.042, and
grassland -0.037. In ``Centro,'' shrubland reaches -0.07, savanna
-0.031, cropland -0.024, forest -0.017, and grassland -0.005 per decade.
This decrease in productivity could be associated either with a
reduction in vegetation surface, a decrease in biomass, or browning
\citep{Miranda2023}.

The temporal variation within the macrozones is showed in
Figure~\ref{fig-zcNDVI_var}b). There is a negative trend in ``Norte
Chico'' with -0.035 and ``Centro'' with -0.02 per decade. Vegetation
reached its lowest values since the year 2019, with an extreme condition
in early 2020 and 2022 in the ``Norte Chico'' and ``Centro''. The
``Sur'' and ``Austral'' show a positive trend of around 0.012 and 0.016,
respectively per decade (Figure~\ref{fig-zcNDVI_var}). Despite the
croplands suffering from drought just as badly as the native vegetation
in ``Norte Chico,'' the savanna and shurbland appears to be the region
most affected by a negative trend in vegetation productivity across
Chile.

@@

Figure~\ref{fig-corTimeScale} is a map that shows the highest
coefficient of determination (r-squared, or rsq) found in the regression
analysis between different drought indicators and plant productivity
over time. The spatial variation of time scales reached per index is
mostly for time scales above 12 months. In the case of SSI, the
predominant scales are 6 and 12 months. For all indices, to the north,
the time scales are higher and diminish toward the south until the south
part of ``Austral'' increases. In Figure~\ref{fig-corPerson}, the map of
Pearson correlation values is shown. The EDDI reached correlations above
0.5 between ``Norte Chico'' and ``Sur.'' The correlation changes from
negative to positive toward the Andes Mountains and to the sea, just as
in the northern part of ``Austral.'' The SPI and SPEI have similar
results, with the higher values in ``Norte Chico'' and ``Centro'' being
higher than 0.6. Following a similar spatial pattern as EDDI. The SSI
showed to be the index that has a major spatial extension with a higher
correlation. It has a similar correlation to SPI and SPEI for ``Norte
Chico'' and ``Sur,''~ but has improvements for ``Sur.''

@@

\hypertarget{drought-trend-and-attribution-to-land-cover}{%
\subsection{Drought trend and attribution to land
cover}\label{drought-trend-and-attribution-to-land-cover}}

\citet{Vicente-Serrano2022}, in a study at the global scale of drought
trends, indicate that there have not been significant trends in
meteorological drought since 1950. Also, state that the increase in
hidrological trend in some parts of the globe (northeast Brazil and the
Mediterranean region) is related to changes in land cover and
specifically to the rapidly increasing irrigated area, which
consequently increases water extraction. \citet{Kogan2020} analyzed the
agricultural drought impact globally and in the main grain producer
countries, finding that \emph{``since 1980, the Earth warming has not
changed the drought area or intensity.''} In our study, we considered
the variation in vegetation productivity in Chile for areas without
changes in land cover, to avoid misleading conclusions that could be
related to the increase in water demand due to land cover change. Our
results show a contrasting perspective. There has been a significant
trend in the decline of vegetation productivity (zcNDVI) since 2000 for
``Norte Chico'' and ``Centro,'' which has been extreme between 2020 and
2022, seemsly due to an intense hydrological drought due to the
persistance of the mega drought \citep{Garreaud2017}. However, a rise in
irrigated land doesn't seem to have an impact on this hydrological
drought. Despite using the persistance mask for vegetation's trend
analysis, cropland, which is the most water-demand type, showed a
decrease trend in ``Norte Chico'' and ``Centro.'' Also, there was an
increase in barren land for both types. These changes are associated
with a decrease in water demand from vegetation. Nonetheless, we used
the persistant land cover to ensure that the pixel has the same class;
in the case of croplands, it could happen that some areas had changed
crops for others with higher water consumption and consequently increase
water demand. But this effect should be minor compared to the results
from land cover change at this scale of analysis.

On the other hand, for ``Norte Chico'' and ``Centro,'' our results show
a decrease in trends of water supply (SPI and SSI), which are higher at
larger time scales, which is evidence of the hydrological drought. We
say that what happened in central Chile goes against what has been found
on a global scale \citep{Vicente-Serrano2022, Kogan2020}. This shows
that the main cause of the hydrological drought in Chile was a steady
drop in water supply made worse by an increase in AED, but it seems that
in zones most affected by drought, the main cause is not an increase in
water demand by vegetation like irrigated crops. Finally, north-central
Chile has experienced a decline in vegetation productivity across all
macroclasses, which is primarily attributable to variations in water
supply and soil moisture. An increase in water demand, such as an
increase in the surface area of irrigated crops, could strengthen this
trend.

\hypertarget{land-cover-types-and-their-impact-by-drought}{%
\subsection{Land cover types and their impact by
drought}\label{land-cover-types-and-their-impact-by-drought}}

>>>>

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover change, vegetation
productivity, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    drought \sep land cover change \sep vegetation productivity \sep 
    satellite
\end{keyword}
\end{frontmatter}
    \captionsetup{justification=raggedright,singlelinecheck=false}

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

For monitoring drought, the World Meteorological Organization recommends
the SPI (Standardized Precipitation Index) \citep{WMO2012}. The SPI is a
multi-scalar drought index that only uses precipitation to assess short-
to long-term droughts. The primary cause of drought is precipitation
anomalies, and temperature usually makes it worse \citep{Luo2017}.
Nowadays, there is an increase in attention toward using AED separately
to monitor droughts \citep{Vicente-Serrano2020}. One reason is due to
its attribution to increasing flash droughts in water-limited regions
\citep{Noguera2022}. \citet{Vicente-Serrano2010} proposed the
Standardized Precipitation Evapotranspiration Index (SPEI), which
incorporated the temperature effect by subtracting AED from
precipitation. SPEI allows for analysis of the combined effect of
precipitation and AED. Since its formulation, it has been used worldwide
for the study and monitoring of drought
\citep{Gebrechorkos2023, Liu2021}. \citet{Hobbins2016} and
\citet{McEvoy2016} developed the Evaporative Demand Drought Index (EDDI)
to monitor droughts solely using the AED, and it has proven effective in
monitoring flash droughts \citep{Li2024, Ford2023}. For soil moisture,
several drought indices exist, such as the Soil Moisture Deficit Index
(SDMI) \citep{Narasimhan2005} and the Soil Moisture Agricultural Drought
Index (SMADI) \citep{Souza2021}. \citet{Hao2013} and
\citet{AghaKouchak2014} proposed the Standardized Soil Moisture Index
(SSI), which has a similar formulation as the SPI, SPEI, and EDDI. Thus,
there are plenty of drought indices that allow for a comprehensive
assessment of drought on short- to long-term scales and that allow for
the use of single variables from the earth's water balance (e.g.,
precipitation, AED, soil moisture). The variation in climate variables
impacts vegetation development, and unfavorable conditions such as low
precipitation and high temperatures usually generate a decrease in
vegetation productivity. To monitor the response of vegetation, the
common practice is to use satellite data. The Normalized Difference
Vegetation Index (NDVI) has been widely used as a proxy for biomass
production \citep{Camps-Valls2021, Paruelo2016, Helman2014}. For Chile's
cultivated land, \citet{Zambrano2018} introduced the zcNDVI for
assessing seasonal biomass production in response to drought. Using this
information, we can advance our understanding of the impact of drought
on ecosystems.

Chile's diverse climatic and ecosystem types
\citep{Beck2023, Luebert2022} make it an ideal natural laboratory for
studying climate and ecosystems. Additionally, the country has
experienced severe drought conditions that have had significant effects
on vegetation and water storage. North-central Chile has faced a
persistent precipitation deficit since 2010, defined as a mega drought.
\citep{Garreaud2017}, which has impacted the Chilean ecosystem. This
megadrought was defined by the Standardized Precipitation Index (SPI) of
twelve months in December having values below one standard deviation.
Some studies have addressed how this drought affects single ecosystems
in terms of forest development \citep{Miranda2020, Venegas2018}, forest
fire occurrence \citep{UrrutiaJalabert2018}, and crop productivity
\citep{Zambrano2023, Zambrano2018, Zambrano2016}. We found one study
regarding land cover and drought in Chile. The study by
\citet{Fuentes2021} evaluates water scarcity and land cover change in
Chile between 29° and 39° of south latitude. \citet{Fuentes2021} used
the SPEI of one month for evaluating drought, which led to misleading
results. For example, they did not find a temporal trend in the SPEI but
found a decreasing trend in water availability and an increase trend on
AED, which in turn should have been capable of being captured with
longer time scales of the SPEI. The term ``megadrought'' in Chile is
used to describe a prolonged water shortage that lasts for several
years, resulting in a permanent deficit that impacts the hydrological
system \citep{Boisier2018}. Therefore, it is crucial to evaluate
temporal scales that consider the cumulative impact over a period of
several years. The association between drought and the environment in
Chile is not well comprehended. Hence, it is imperative to acquire a
more profound comprehension of the manner in which climatic and soil
moisture droughts influence environmental dynamics, in order to make
well-informed decisions on adaptation strategies.

@@

\hypertarget{short--to-long-term-drought}{%
\subsection{Short- to long-term
drought}\label{short--to-long-term-drought}}

@@

We analyzed the trend of vegetation productivity over the unchanged land
cover macroclasses. To achieve this, we used the persistent mask of land
cover macroclasses. This way, we tried to reduce the noise in the
vegetation due to a change in land cover from year to year. We used the
zcNDVI as a proxy of vegetation productivity. In Chile's cultivated
land, \citet{Zambrano2018} introduced the zcNDVI for assessing seasonal
biomass production in relation to climate.

We examine the drought indices of water demand, water supply, and soil
moisture and their correlation with vegetation productivity. The
objective is to determine the impact of soil moisture and water demand
and supply on vegetation productivity. We want to address three main
questions: Which of the drought variables---supply, demand, or soil
moisture---most helps to explain the changes in plant productivity? How
do the short- to long-term time scales of the drought variable affect
vegetation productivity in Chile? And finally, how strong is the
relationship between the variables and the drought index? Thus, we will
be able to advance in understanding how climate is affecting vegetation,
considering the impact on the five land cover types: forest, cropland,
grassland, savanna, and shrubland.

We conducted an analysis on the linear correlation between the indices
SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36
months with zcNDVI. We used a method similar to that used by
\citet{Meroni2017} which compared the SPI time-scales with the
cumulative FAPAR (Fraction of Absorbed Photosynthetically Active
Radiation). We performed a pixel-to-pixel linear correlation analysis
for each index within the persistent mask of land cover macroclasses. We
first compute the Pearson coefficient of correlation for each of the six
time scales. A time scale is identified as the one that attains the
highest correlation (p \textless{} 0.05). We then extracted the Pearson
correlation coefficient corresponding to the time scales where the value
peaked. As a result, for each index, we generated two raster maps: 1)
containing the raster with values of the time scales and drought index
that reached the maximum correlation, and 2) having the magnitude of the
correlation obtained by the drought index at the time scales.

@@

For the downloading, processing, and analysis of the spatio-temporal
data, we used the open source software for statistical computing and
graphics, \texttt{R} \citep{R2023}. For downloading ERA5L, we used the
\texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing raster
data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
used \texttt{\{SPEI\}} \citep{Bergueria2023}. For mapping, we use
\{tmap\} \citep{Tennekes2018}. For data analysis and visualization, the
suite \{tidyverse\} \citep{Wickham2019} was used. For the random forest
modeling, we used the \{tidymodels\}\citep{Kuhn2020} and
\{ranger\}\citep{Wright2017} packages.

@@

\hypertarget{short--to-long-term-drought-1}{%
\subsection{Short- to long-term
drought}\label{short--to-long-term-drought-1}}

@@

The Figure~\ref{fig-trendDIMacro} displays the averaged aggregation per
macrozone, the drought index, and the timescale. The macrozones that
reached the lowest trend for SPI, SPEI, and SSI are ``Norte Chico'' and
``Centro,'' where the indices also decrease at longer time scales.
Potentially explained due to the prolonged reduction in precipitation
that has affected the hydrological system in Chile. At 36 months, it
reaches trends between -0.03 and -0.04 (z-score) per decade for SPI,
SPEI, and SSI. For ``Sur,'' the behavior is similar, decreasing at
longer scales and having between -0.016 and -0.025 per decade for SPI,
SPEI, and SSI. ``Norte Grande'' has the highest trend at 36 months for
EDDI (0.042 per decade), and ``Centro'' has the lowest for SPI and SPEI.
In ``Norte Grande'' and ``Norte Chico,'' which are in a semi-arid
climate, it is evident that the EDDI has an effect on the difference
between the SPI and SPEI index, which is not seen in the other
macrozones. Contrary to the other macrozones, ``Austral'' showed an
increase in all indices, being the highest for EDDI at 36 months (0.025)
and the lowest for SSI, which shows only a minor increase in the trend.

\hypertarget{interaction-of-land-cover-and-drought-1}{%
\subsection{Interaction of land cover and
drought}\label{interaction-of-land-cover-and-drought-1}}

\hypertarget{land-cover-change-1}{%
\subsubsection{Land cover change}\label{land-cover-change-1}}

@@

From the trend analysis in Table \ref{tab-landcoverTrend}, we can
indicate that the ``Norte Chico'' shows an increase in barren land of
111 \(km^2 yr^{-1}\) and a reduction in the class savanna of 70
\(km^2 yr^{-1}\). In the ``Centro'' and ``Sur,'' there are changes with
an important reduction in savanna (136 to 318 \(km^2 yr^{-1}\) ), and an
increase in shrubland and grassland. Showing a change for more dense
vegetation types. The area under cultivation (croplands) appears to be
shifting from the ``Centro'' to the ``Sur.'' Also, there is a high
increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' seemingly
replacing the savanna lost (Table \ref{tab-landcoverTrend}).

@@

The temporal variation within the macrozones is shown in
Figure~\ref{fig-zcNDVI_var}b). There is a negative trend in ``Norte
Chico'' with -0.035 and ``Centro'' with -0.02 per decade. Vegetation
reached its lowest values for 2019-2022, with an extreme condition in
early 2020 and 2022 in the ``Norte Chico'' and ``Centro''. The ``Sur''
and ``Austral'' show a positive trend of around 0.012 and 0.016,
respectively, per decade (Figure~\ref{fig-zcNDVI_var}).

In Figure~\ref{fig-zcNDVI_var} it is showed the spatial map of trends in
zcNDVI (Figure~\ref{fig-zcNDVI_var}a). In ``Norte Grande,'' vegetation
productivity, as per the z-index, exhibits a yearly increase of 0.027
for grassland and 0.032 for shrubland. In the ``Norte Chico,'' savanna
has the lowest trend of -0.062, cropland -0.047, shrubland -0.042, and
grassland -0.037. In ``Centro,'' shrubland reaches -0.07, savanna
-0.031, cropland -0.024, forest -0.017, and grassland -0.005 per decade.
This decrease in productivity could be associated either with a
reduction in vegetation surface, a decrease in biomass, or browning.

@@

Figure~\ref{fig-corTimeScale} shows the highest coefficient of
determination (r-squared, or rsq) found in the regression analysis
between different drought indicators and zcNDVI over time scales (1, 3,
6, 12, 24, and 36 months). The spatial variation of time scales reached
per index is mostly for time scales above 12 months. In the case of SSI,
the predominant scales are 6 and 12 months. For all indices, to the
north, the time scales are higher and diminish toward the south until
the south part of ``Austral,'' where they increase. In
Figure~\ref{fig-corPerson}, the map of Pearson correlation values (r) is
shown. The EDDI reached correlations above 0.5 between ``Norte Chico''
and ``Sur.'' The correlation changes from negative to positive toward
the Andes Mountains and to the sea, just as in the northern part of
``Austral.'' The SPI and SPEI have similar results, with the higher
values in ``Norte Chico'' and ``Centro'' being higher than 0.6.
Following a similar spatial pattern as EDDI but with an opposite sign.
The SSI showed to be the index that has a major spatial extension with a
higher correlation. It has a similar correlation to SPI and SPEI for
``Norte Chico'' and ``Sur,''~ but has improvements for ``Sur.''

@@

\hypertarget{drought-trend-and-attribution-to-land-cover-change}{%
\subsection{Drought trend and attribution to land cover
change}\label{drought-trend-and-attribution-to-land-cover-change}}

\citet{Vicente-Serrano2022}, in a study at the global scale of drought
trends, indicates that there have not been significant trends in
meteorological drought since 1950. Also, state that the increase in
hidrological trend in some parts of the globe (northeast Brazil and the
Mediterranean region) is related to changes in land cover and
specifically to the rapidly increasing irrigated area, which
consequently increases water extraction. \citet{Kogan2020} analyzed the
agricultural drought impact globally and in the main grain producer
countries, finding that ``since 1980, the Earth warming has not changed
the drought area or intensity.'' In our study, we took into account the
variation in vegetation productivity in Chile, specifically in areas
without any changes in land cover, to prevent any misleading conclusions
about the increase in water demand due to land cover change. Our results
show a contrasting perspective. The SPI, SPEI, and SSI (water supply)
showed a decrease in trends, except for the southern part, and an
increase in EDDI (water demand). The trend, positive or negative, was
stronger as the time scales increased. Trends in the long term (e.g., 36
months) are evidence of how human-induced climate change is affecting
Chile, which seems to be due to an intense hydrological drought
resulting from the persistence of the precipitation deficit. We found
that there has been a significant trend in the decline of vegetation
productivity (zcNDVI) since 2000 for the north-central part of the
country, which has been extreme between 2020 and 2022 and has impacted
natural and cultivated land. Additionally, we demonstrated that the
drought, primarily due to an increase in AED, accounts for about 30\% of
the changes in land cover types (excluding croplands). These changes are
associated with a decrease in water demand from vegetation. Moreover,
the most water-demanding type, cropland, showed a decrease in the
north-central region, while barren land showed an increase. The
north-central part of the country primarily experienced these changes
due to a higher increase in AED. Thus, we have evidence of a significant
decline in water supply and an increase in AED for the north-central
part of Chile, which show to be the most relevant variables for drought
conditions. Some questions arise regarding what is occurring with the
cultivated land. We used the unchanged land cover to ensure that an
increase in surface area is not considered in the trend analysis. For
croplands, it could happen that some areas have changed the types of
crops for others with higher water demand, which consequently increases
water demand. However, this effect should be minor compared to the
decrease in water supply and increase in water demand at this scale of
analysis.

This shows that the main cause of the hydrological drought in Chile was
a steady drop in water supply made worse by an increase in AED, but it
seems that in zones most affected by drought, the main cause is not an
increase in water demand due to an intensification of cultivated land
(e.g., an increase in irrigated crops). North-central Chile has
experienced a decline in vegetation productivity across land cover
types, which is primarily attributable to variations in water supply and
soil moisture. An increase in water demand, such as an increase in the
surface area of irrigated crops, could strengthen this trend.

\hypertarget{impact-of-short--to-long-term-droughts-on-land-cover-types}{%
\subsection{Impact of short- to long-term droughts on land cover
types}\label{impact-of-short--to-long-term-droughts-on-land-cover-types}}
